\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-16 chapters/patients.tex
BCI chapter

\section{Patient presentation}
\section{Data collection \& preprocessing}
\todo{Define structure patient results/discussion}
\section{Outcomes}
\section{Discussion}

>>>>

\section{Introduction}

Communication BCIs have a target population consisting of patients in various
stages of paralysis or Locked-in Syndrome, often suffering from oculomotor im-
pairments. This reduces their performance operating visual event-related poten-
tial (ERP) BCIs, since they are unable to comfortably redirect their gaze at the de-
sired target, i.e., overt visuospatial attention (VSA). Instead, they are forced to op-
erate in covert VSA, where the gaze and VSA do not coincide. Several studies show
that performance drops in covert VSA [2, 1], necessitating gaze-independent so-
lutions. Our previous study [3] indicates gaze-independent performance can be
improved in healthy subjects using a suited decoding strategy. Currently, we aim
to verify this in patients.

\section{Materials \& methods}
\subsection{Patient recruitment}
\subsection{Eye motor examination}
\todo{include eye test document juliette}
\subsection{Data collection \& preprocessing}
\subsection{BCI Decoding}
\section{Patient presentation}
\todo{Define structure patient results/discussion}
\section{Outcomes}
\section{Discussion}

### 2024-09-16 chapters/problem_statement.tex
BCI chapter

\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}

\todo{Check provisionary and final doctoral plan for motivation}
\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities, by leveraging a non-invasive high-ITR
visuospatial ERP-based BCI and improving ERP decoding performance (in general
and specifically in gaze-independent settings).}
\todo{check WIP seminar}

\todo{update text}

\emph{Novelty:}

Brain-Computer Interfaces (BCIs) decode brain activity with the aim to establish a direct communication
pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
devoid of these abilities, for whom BCIs can provide a means to communicate or to
control devices.
The quest for performant and affordable solutions is most evident in
visual BCIs based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
and paradigms tailored to the needs of specific stimulation paradigms and
use contexts.
An effective and proven method is the visual event-related potential (ERP)-based
oddball interface.
Here, targets are shown in short pulses on a computer screen.
Each time the user observes a target, an ERP is evoked, the presence of which can be
detected in the EEG-signal.
The ERP consists of multiple components, of which some are modulated by the
attention of the user, specifically the N2 and the P3 component.
Decoding this modulation allows for information transfer controlled by the
user's brain activity, as sketched in Figure~\ref{fig:bci}.
However, visual BCIs cannot operate efficiently when the user does not direct
their gaze onto the
desired target~\cite{Brunner2010}.

This leads to an often unadressed paradox: the BCI target population consists
of patients who suffer from pathologies like Amyotrophic Lateral Sclerosis
(ALS), Multiple Sclerosis (MS), stroke (specifically
brain-stem stroke) or traumatic brain injury, which can lead to several degrees
of paralysis, or even to Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While BCIs are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking, studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Table~\ref{tab:incidence} reports the relatively high frequencies of
eye motor problems, which can range from involuntary eye movements to control
issues and even to partial or complete ophtalmoplegia.
\begin{table}
	\centering
	\begin{tabular}{@{}|l|rrrrr|r|@{}}
		\hline
		         & ALS  & MS   & Stroke  &
		DMD      & SMA  & LiS                           \\ \hline
		Minor    & 50\% & 31\% & 40-70\% & + & - &      \\
		Severe   & 33\% & 3\%  & +       & - & - & 98\% \\
		Complete & 17\% & -    & +       & - & - & 2\%  \\
		\hline
	\end{tabular}
	\caption{Incidence of eye motor impairment in Brain-Computer Interface target
		populations: (ALS: Amyotrophic Lateral Sclerosis, MS: Multiple Sclerosis,
		DMD: Duchenne's Muscular Dystrophy, SMA: Spinal Muscular Atrophy, LiS:
		Locked-In-Syndrome).}\label{tab:incidence}
\end{table}
\todo{document methodology}
\todo{find more references, fill out}
Different degrees of eye motor impairment can render visual BCIs
uncomfortable or outright impossible to use.
When operating a visual BCI, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the BCI.
This points to the need for gaze-independent BCIs.

@@

The general goal of this PhD is to tackle the gaze-dependency in visual BCIs.
Traditionally, gaze-independency~\cite{Riccio2012, Aloise2012} can be realized in three ways:
Firstly, by avoiding visual
stimulation entirely and opting for auditory or somatosensory stimulation instead.
However, these alternatives often result in lower information transfer rates,
increased mental effort and variable outcomes for different users.
Secondly, gaze-independence can
be achieved by adapting the interface to display stimuli such that they
are always in the center of the field of view, by exploiting visual
non-spatial attention (feature attention), or a combination thereof.
These interfaces suffer equally from reduced information transfer rates.
Moreover, they still rely to some extent on eye motor control, necessitating central gaze fixation.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like Rapid Serial Visual Presentation,
where targets or small sets of targets are shown consecutively.
Traditionally, spatially organized interfaces have a higher information transfer
rate, but are also more gaze-dependent than temporally organized interfaces.
Third, stimuli can be presented in a standard BCI paradigm, but \emph{mental} attention
can be decoded separately from \emph{visual} attention.
\todo{table with paradigms and itrs from WIP seminar}

This leads us to adopt a specific approach: improve the performance of a
spatially organized visual oddball ERP-based brain
computer interface by using a suited decoding strategy.
Hence, our working hypothesis is as follows:
\begin{quote}
	\textit{A spatially organized visual oddball ERP BCI with a suited decoder
		can achieve similar or higher information transfer rate in patient
		populations with eye motor deficits than other
		gaze-independent alternatives described in literature.}
\end{quote}

@@

\section{Methods}
\todo{update with new decoders and experiments}

\subsection{Interface design}
We designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
decoders and on our own proposed decoders.
In a first series of experiments, we recorded data with this interface from healthy participants.
The goal of these experiments is to benchmark
gaze-independent ERP decoding algorithms.
We denote this dataset as the Covert Visuospatial Attention ERP dataset
(CVSA-ERP)
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).

Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG and electro-oculogram (EOG) are
recorded, as well as eye gaze using eye tracking.
To simulate the dissociation between the eye gaze (visual attention) and the
intended target (mental attention) which could occur in patients with hampered eye
motor control, healthy participants are cued to operate the BCI in specific
visuospatial attention (VSA) conditions.
Each subject performs 5 different VSA conditions as illustrated in
Figure~\ref{fig:vsa-conditions}.
\begin{figure}
	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
		asked to dissociate their visual and mental attention by gazing at a
		crosshair while attending a target, depending on the prompted condition.}%
	\label{fig:vsa-conditions}
\end{figure}
\todo{new picture that shows more clearly a user operating freely where they
can gaze}
In the overt case, subjects are instructed to gaze at the cued
target they are also mentally attending; in the covert case, subjects are instructed
to gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent BCI development: split VSA.
In split VSA, the participant mentally focuses on one cued target while gazing at
another (the distractor).
Depending on the inter-target distance along the hull of the hexagon between the
attended target and the distractor we discern three split VSA subconditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite the intended target
($d=3$).

@@

\todo{update with current results and all chapters}

>>>>

\todo{provisionary doctoral plan}
\todo{check WIP seminar}
\todo{update text}

@@

Brain-Computer Interfaces (BCIs) decode brain activity with the aim to establish a direct communication
pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
devoid of these abilities, for whom BCIs can provide a means to communicate or to
control devices.
The quest for performant and affordable solutions is most evident in
visual BCIs based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
and paradigms tailored to the needs of specific stimulation paradigms and
use contexts.
An effective and proven method is the visual event-related potential (ERP)-based
oddball interface.
Here, targets are shown in short pulses on a computer screen.
Each time the user observes a target, an ERP is evoked, the presence of which can be
detected in the EEG-signal.
The ERP consists of multiple components, of which some are modulated by the
attention of the user, specifically the N2 and the P3 component.
Decoding this modulation allows for information transfer controlled by the
user's brain activity, as sketched in Figure~\ref{fig:bci}.
However, visual BCIs cannot operate efficiently when the user does not direct
their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.

This leads to an often unadressed paradox: the BCI target population consists
of patients who suffer from pathologies like Amyotrophic Lateral Sclerosis
(ALS), Multiple Sclerosis (MS), stroke (specifically
brain-stem stroke) or traumatic brain injury, which can lead to several degrees
of paralysis, or even to Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While BCIs are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking, studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Different degrees of eye motor impairment can render visual BCIs
uncomfortable or outright impossible to use.
When operating a visual BCI, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the BCI.
This points to the need for gaze-independent BCIs.

The general goal of this PhD is to tackle the gaze-dependence in visual BCIs.
\todo{refer to ITRs from previous chapter}
Section{sec:gaze-independence-paradigms} showed that visual BCI paradigms form
a good candidate for a performant gaze-independent BCI.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like Rapid Serial Visual Presentation,
where targets or small sets of targets are shown consecutively.
Traditionally, spatially organized interfaces have a higher information transfer
rate, but are also more gaze-dependent than temporally organized interfaces.
In such an interface, stimuli can be presented in a standard BCI paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

This leads us to adopt a specific approach: improve the performance of a
spatially organized visual oddball ERP-based brain
computer interface by using a suited decoding strategy.
Hence, our working hypothesis is as follows:
\begin{quote}
	\textit{A spatially organized visual oddball ERP BCI with a suited decoder
		can achieve similar or higher information transfer rate in patient
		populations with eye motor deficits than other
		gaze-independent alternatives described in literature.}
\end{quote}
\todo{tone this down a bit to what can be proven}

@@

\todo{\emph{Goal:} Enable communication for eye-motor impaired patients}
\todo{\emph{Method:} Design a comfortable interface that allows them to maximally exploit
their residual gaze capabilities, by leveraging a non-invasive high-ITR
visuospatial ERP-based BCI and improving ERP decoding performance (in general
and specifically in gaze-independent settings).}

\section{Methods}

@@

\todo{Tensor discriminant analysis}

\subsection{Interface design}

\todo{Overt, covert, split, free, illustration from poster}
\todo{clear definitions}

\subsection{Experiments}
\subsubsection{Healthy participant study}
We designed a visual oddball interface that helps us study and adapt
to the effects of eye motor impairment on the ERP, when using existing state-of-the-art
decoders and on our own proposed decoders.
In a first series of experiments, we recorded data with this interface from healthy participants.
The goal of these experiments is to benchmark
gaze-independent ERP decoding algorithms.
We denote this dataset as the Covert Visuospatial Attention ERP dataset
(CVSA-ERP)
This study was approved by the Ethics Commission of University Hospital Leuven
(S62547).

Using a hexagonal lay-out interface, similar to the visual Hex-o-Spell proposed
by Treder \& Blankertz~\cite{Treder2010}, we present six flashing targets (without letters or
symbols) to the participant while the EEG and electro-oculogram (EOG) are
recorded, as well as eye gaze using eye tracking.
To simulate the dissociation between the eye gaze (visual attention) and the
intended target (mental attention) which could occur in patients with hampered eye
motor control, healthy participants are cued to operate the BCI in specific
visuospatial attention (VSA) conditions.
Each subject performs 5 different VSA conditions as illustrated in
Figure~\ref{fig:vsa-conditions}.
\todo{do we already want this here?}
\begin{figure}
	%\includegraphics[width=\linewidth]{figures/attention_modes.eps}
	\caption{Visuospatial attention (VSA) conditions in our experiment. The user is
		asked to dissociate their visual and mental attention by gazing at a
		crosshair while attending a target, depending on the prompted condition.}%
	\label{fig:vsa-conditions}
\end{figure}
\todo{new picture that shows more clearly a user operating freely where they
can gaze}
In the overt case, subjects are instructed to gaze at the cued
target they are also mentally attending; in the covert case, subjects are instructed
to gaze at the center of the screen while mentally attending to the cued target.
Finally, we introduce a VSA condition that is understudied in the context of
gaze-independent BCI development: split VSA.
In split VSA, the participant mentally focuses on one cued target while gazing at
another (the distractor).
Depending on the inter-target distance along the hull of the hexagon between the
attended target and the distractor we discern three split VSA subconditions:
the distractor is either clockwise or counterclockwise directly next to the
attended target ($d=1$), there is one other target between the attended target and
the distractor ($d=2$), or the distractor is opposite the intended target
($d=3$).

\subsubsection{Patient study}
\todo{}

### 2024-09-16 chapters/stbf_struct.tex
Add wcble paper
BCI chapter

\newcommand{\pv}[1]{
	%\ifdim#1pt<0.05pt
	%	\cellcolor{black!20}
	%\fi
	%\ifdim#1pt<0.001pt
	%	$< 0.001$
	%\else
	%	\ifdim#1pt=1.000pt
	%		\multicolumn{1}{c}{--}
	%	\else
	%		$#1$
	%	\fi
	%\fi
	#1
}
% Kronecker-structured LDA ====================================================
\begin{refsection}
	\chapter{Kronecker-structured discriminant analysis}

@@

  \todo{Include part kronlda, check notes in journal}
\section{Introduction}

LCMV struct might be to restrictive
Toeplitzlda
MEG paper

\section{Methods}
\subsection{Spatiotemporal Linear Discriminant Analysis}
\subsection{The single Kronecker product covariance model}
\begin{equation}
	\mat{C} = \mat{S}\otimes\mat{T}
\end{equation}
Tapering, Toeplitz and shrinkage
\subsection{The Kronecker sum covariance model}
\begin{equation}
	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
\end{equation}
pvl, kronpca

\subsection{Kronecker sum covariance estimation}
\subsection{Classification}
\section{Results}
\section{Discussion}

\todo{check chapters/structure paper}
\todo{check structure Leuven.ai poster}
\section{The spatiotemporal EEG covariance}
\section{The single-KP covariance model}
\section{The sum-of-KP covariance model}
\section{Decoding performance}
\section{Extension into the space-time-frequency domain}

	\todo{check chapters/structure paper}
	\todo{check structure Leuven.ai poster}
	\section{The spatiotemporal EEG covariance}
	\section{The single-KP covariance model}
	\section{The sum-of-KP covariance model}
	\section{Decoding performance}
	\subsection{Extension into the space-time-frequency domain}
	\printbibliography
\end{refsection}

>>>>

\newcommand{\pv}[1]{
	%\ifdim#1pt<0.05pt
	%	\cellcolor{black!20}
	%\fi
	%\ifdim#1pt<0.001pt
	%	$< 0.001$
	%\else
	%	\ifdim#1pt=1.000pt
	%		\multicolumn{1}{c}{--}
	%	\else
	%		$#1$
	%	\fi
	%\fi
	#1
}
% Kronecker-structured LDA ====================================================

@@

  \todo{Include part kronlda, check notes in journal}
\todo{
LCMV struct might be to restrictive
Toeplitzlda
MEG paper}
%\begin{equation}
%	\mat{C} = \mat{S}\otimes\mat{T}
%\end{equation}
%Tapering, Toeplitz and shrinkage
%\subsection{The Kronecker sum covariance model}
%\begin{equation}
%	\mat{C} = \sum_k^{CS}\mat{S}\otimes\mat{T}
%\end{equation}
\todo{pvl, kronpca}

\todo{check chapters/structure paper}
\todo{check structure Leuven.ai poster}
\todo{extension to the TFR domain}

### 2024-09-16 chapters/wcble.tex
BCI chapter

\chapter{Woody Classifier-based Latency Estimation}
\section{Latency estimation and correction}
\todo{check earlier draft alignment paper}
\todo{check deleted paragraphs in jitter paper}
\section{Classifier-based Latency Estimation}
\section{Classifier-based Latency Estimation with Woody iterations}
\section{Simulation study}

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/wcble/figure1.pdf}
	\caption{Schematic representation of the Woody Classifier Based Latency

@@

		classifier.}
	\label{fig:diagram-train}
\end{figure*}

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/wcble/figure2.pdf}
	\caption{Schematic representation of the Woody Classifier Based Latency

@@

		trained second-stage classifier, which predicts the label of the epochs.}
	\label{fig:diagram-eval}
\end{figure*}

@@

\todo{insert comments}

>>>>

\section{Introduction}

\todo{General introduction of why we're doing this}

\subsection{ERP latency jitter}

\subsubsection{Neural origins of latency jitter in the time domain}
While the classic ERP model assumes a constant ERP response over trials,
research has shown that cortical evoked potentials show significant
trial-to-trial jitter~\cite{truccolo2002trial}.
Experiments show that latency jitter in ERPs is the result of the interacion of
the evoked activity with ongoing dynamics in the brain~\cite{Hasenstaub2007,
	kisley1999trial, Curto2009, Arieli1996}.

The latency of an evoked potential is a combination of the modulation of oscillating activity that is
time-locked, but not necessarily phase-locked to stimulus onset, and a
delay caused by ongoing neurophysiological processes that are not necessarily related to the task.
\todo{Check this, \cite{Stokes2016,Mouraux2008}}
\todo{Read references and check if neural origins and
	generation model justify approach}

\subsubsection{Implications for ERP analysis}
The presence of latency jitter introduces multiple issues to ERP analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:smearing}. In ERP analysis, single-trial ERPs are usually
averaged per condition to cope with the unfavorable signal-to-noise ratio  When
averaging over multiple identical signals with jittered latencies, the shape and amplitude of the average do not reflect the properties
of the original signal. Because the amplitude of the average is lower, the
signal-to-noise ratio will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting ERP
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{woody1967characterization, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

\todo{illustrate smearing effect}
%\begin{figure}
%  \todo{smearing effect}
%	\caption{The smearing effect: averaging over jittered waveforms leads to a spread out
%		average with a lower amplitude.}
%	\label{fig:smearing}
%\end{figure}

\subsubsection{Benefits of latency estimation}
A common approach to counteract the smearing effect is correcting single-trial
ERPs for jitter before averaging.\todo{reference} This can be done by estimating the latency of each
single-trial ERP, and aligning them before averaging.
Latency estimation has multiple advantages:
\begin{enumerate*}[label={(\arabic*)}]
	\item Latency-based ERP alignment can help reveal the true shape and
	properties like peak amplitude and latency of ERPs across
	experimental conditions. These are variables of interest for
	testing electrophysiological an behavioral hypotheses.
	Additionally, latency estimation allows for the extraction of
	other descriptors from a set of single-trial ERPs, like
	amplidute or latency variance~\cite{Hultsch2004}, which can
	also be of interest as studied variables. For instance, in
	\cite{Saville2014}, the inter-trial latency variability itself
	is determined as a correlate of working memory performance.
	\item Counteracting the smearing effect will boost the signal-to-noise
	ratio of averaged ERPs by increasing the amplitude of the
	ERP signal. This in turn can reveal smaller or more jittered ERP components
	which are otherwise obscured or blurred due to desynchronization.
	These deblurred templates can then be used for analysis of the ERP
	waveform, or as templates of the signal of interest in
	BCIs~\cite{arico2014influence}.
	%this enables decoders to take into account the relatively low amplitude
	%earlier components, leading to applications in settings where
	%information from other components than the P300 need to be exploited,
	%like covert visual attention paradigms~\cite{arico2014influence,
	%hardiansyah2020single}.\todo[inline]{rework this to latency features
	%because smaller visual components might not really be revealed
	%since they're already time-locked}
	\item Finally, the introduction of latency estimation and alignment generaly
	improves BCI classification performance. Single-trial latencies can be
	used as features to improve classification~\cite{hardiansyah2020single}.
	Furthermore, compensating for inter-trial differences, especially when
	extended to inter-subject differences, can aid in transfer learning or
	generalization across subjects and protocols\cite{Iturrate2014}.
\end{enumerate*}
\todo{Read, summarize and cite BCI	references from bibliography}

\subsection{Existing methods for latency estimation and alignment}

\subsubsection{Single-component approaches}
The most straight-forward method to estimate single-trial latency
peak-picking. Here, the latency of an ERP is determined as the
time-point of it's maximum (or minimum) amplitude with relaion to the stimulus
onset. While straight-forward, this method
does not perform well in low or negative SNR conditions, unless combined with
filtering. Filtering can suppress the noise contaminating EEG trials, lowering
the risk of picking a noise peak instead of a true peak of the ERP. Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{Carlton1980, McCarthy1981,
	Karjalainen1999, Nishida1999,  Sparacino2002, Georgiadis2005,
	D’Avanzo2011} and in the time-frequency
domain~\cite{Quiroga2003, wang2007single,Hu2010,arico2014influence}. Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using PCA~\cite{Ouyang2017}, ICA~\cite{townsend1999analyzing, jung2001analysis,
	Milne2011}, or as a spatial LCMV beamformer~\cite{TREDER2016279}.

A generally more preferred approach than
peak-picking is template matching. Woody's algorithm~\cite{woody1967characterization} is a simple and performant latency
estimation method that is still being used and is considered amongst the more
performant techniques~\cite{Ouyang2017}. It starts with the average ERP as a
rough estimation of the aligned template ERP, and iteratively refines this
template by determining at each step the time point of maximum
cross-correlation of this template with all single-trial ERPs and aligning them
on these time points to form the next template.
Similarily, cross-correlation template matching can be combined with filtering.
\cite{souloumiac2013improved} uses XDAWN to construct spatial filter before
template matching the spatially filtered signal. \cite{Iturrate2014} uses a
spatiotemporal filter. Instead of aligning the
latencies to obtain the template for the next iteration, weighted averaging can
also be used~\cite{gasser1983selavco}.

Despite good performance, the iterative scheme of Woody's algorithm has a risk of converging to a local
optimum. To avoid this, a fitness function can be defined to jointly optimize
the set of single-trial ERP latencies, which can be optimized with a genetic algorithm~\cite{Pelo2018}.

Methods adapted to handle multiple conditions, useful in BCI setting
\begin{itemize}
	\item \cite{Kim2020}
\end{itemize}

\subsubsection{Multi-component approaches}
The abovementioned methods do not take into account the fact that  ERPs are multi-component signals. Multiple components can be time-locked to different events or neurophysiological proccess and hence are not necessarily
time-locked to each other. They each can have different latency
variabilities. The presence of more than one component can hamper the
performance of some algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple ERP components
by carefully selecting peaks or pre-determined regions of interest.
\cite{hardiansyah2020single} uses an SNR boosting method, applied to a specific
time-window for each ERP component to extract a latency per component.
Methods based on spatial decomposition filters like PCA, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
ERP components. A clear evaluation is lacking of the performance of these
methods in
multi-component settings is lacking.

Some algorithms separate components into stimulus and response locked component clusters based on their
latency distributions~\cite{jung2001analysis, Takeda2008,
	Zhang1998, Yin2009}. These algorithms do, however, require a response
button trigger signal as input to determine the response time, which is
not always present or applicable in an ERP experiment, especially in
the case of BCIs.
The RIDE algorithm~\cite{Ouyang2011, Ouyang2015,
	Wang2015, Ouyang2016, Ouyang2020} is able to do this without entirely depending on
the response time, and can separate stimulus-locked, response-locked,
and multiple central component clusters. Others, like spatial filtering methods that use
decomposition methods to determine the spatial filter can be be modified to separate
components.~\cite{Ouyang2017}.

To categorize:\todo{categorize and analyze these references}
\begin{itemize}
	\item \cite{Dimitriadis2018}
	\item \cite{Thornton2008}
	\item Maximum likelihood~\cite{tuan1987variable, Moecks1988,Puce1994} Maximum likelihood
	\item \cite{Gratton1989} (no access) Maximum likelihood?
\end{itemize}

\todo{Explain beamforming}

Finally, the smearing problem can also be overcome by algorithms that are not
based on ERP component latency estimation, but which use time warping-based methods
to enhance the shape of the averaged template ERP, like Dynamic Time
Warping~\cite{486255, wang2001warp, Zoumpoulaki2015}, Correlation-Optimized
Warping~\cite{Skov2006}, or Fast Variational
Alignment~\cite{flotho2021fast}. Since these methods work
fundamentally different from the latency estimation methods, and cannot
directly be used to extract latency features, we will not further consider them
in this work.

\subsubsection{Contribution}
In this work, we introduce a new latency method that leverages data-driven,
joint spatiotemporal filtering based on the LCMV beamformer~\cite{Vliet2016}. The algorithm
integrates the spatiotemporal filtering in Woody's template matching iteration
scheme and can easily be extended to multi-component settings.
Our method attempts to model the spatial and temporal activation of
each ERP component, as well as it's latency distribution, and takes into acount the
spatial and temporal correlation of the EEG background noise.

@@

%\begin{figure}
%	\includegraphics[width=\linewidth]{figures/wcble/figure1.pdf}
%	\caption{Schematic representation of the Woody Classifier Based Latency
%    Estimation training phase.
%    (1) The first-stage spatiotemporal binary classifier is trained on a set of
%    epochs.
%    (2) It is	then applied to time-shifted copies of the epochs to obtain scores
%    and class probabilities	over time.
%    (3) The medians of these probability distributions are assumed the class
%    latencies.
%    (4) The epochs are aligned to their corresponding	class latencies by
%    shifting in time such that all latencies fall at the same moment.
%    (5) The spatiotemporal classifier is then retrained on the aligned epochs
%    for a	next iteration.
%    (6) After the iterative process halts, the scores and	latencies obtained
%    from the last iteration are used to train the second-stage
%		classifier.}
%	\label{fig:diagram-train}
%\end{figure}
\fullpagefig{figures/wcble/figure1.pdf}{Schematic representation of the Woody Classifier Based Latency

@@

    classifier.}{fig:diagram-train}

@@

\fullpagefig{figures/wcble/figure2.pdf}{Schematic representation of the Woody Classifier Based Latency

@@

    trained second-stage classifier, which predicts the label of the epochs.}{fig:diagram-eval}

@@

\todo{Pr not italic}
\todo{re-insert comments}
